EuPPollNet: A European database of plant-pollinator networks

### 2024-06-03 Manuscript/EuPPollNet_word.qmd
edits from co-authors

**KEYWORDS**\
\begingroup
\small
plant-pollinator networks, plant-pollinator interactions, flowering plants, Angiosperms, pollinators, nestedness, connectance
\endgroup
\pagebreak

@@

The interaction between plants and pollinators is one of the most well-documented mutualisms on Earth. Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutionary history [@ollerton2017], and are critically important for terrestrial biodiversity and the human economy. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species and guiding conservation planning [@heberling2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which can impact the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016]. Further, despite little sampling effort could capture most of the relevant functional species [@hegland2010], most plant-pollinator networks have unobserved interactions because undersampling [@olesen2011; @chacoff2012]. Thus, the lack of strong spatio-temporal coverage, along with the current intrinsic limitations of sampling plant-pollinator networks, highlight the need of keep gathering and integrating informative species interaction data to properly unravel the different ecological processes that shape plant-pollinator interactions at large scales.

@@

Bees conducted the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were especially fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only species rich in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. Honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted in average a third of the total interactions per network. Indeed, this proportion is higher that the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights their key role as potential pollinator of many flowering plants. However, high honey bee densities can have negative impacts on wild pollinators and the plants they visit [@steffan2000; @geldmann2018; @herrera2020], emphasising the need to evaluate these impacts and implement actions if necessary.

>>>>

**KEYWORDS**\
\begingroup
\small
plant-pollinator networks, species interactions, flowering plants, Angiosperms, pollinators, nestedness, connectance
\endgroup
\pagebreak

@@

The interaction between plants and pollinators is one of the most well-documented mutualisms on Earth. Plant-pollinator interactions involve a great diversity of species, largely attributed to their coevolutionary history [@ollerton2017], and are critically important for terrestrial biodiversity and economic productivity. The synergistic effects of climate change with other global change pressures are threatening worldwide biodiversity [@sala2000; @bellard2014], including plant and pollinator species as well as their interactions [@goulson2015; @settele2016; @eichenberg2021]. Under this scenario, the increasing availability of biodiversity data plays a major role in our ecological understanding of species status, trends, and conservation [@heberling2021; @zattara2021]. However, our knowledge of plant and pollinator species and their network of interactions still exhibits major temporal, spatial and taxonomic biases [@archer2014; @troia2016; @poisot2021; @marshall2024], limiting our ability to effectively protect their biodiversity.

@@

The interactions between different plant and pollinator species within a community form complex networks. Macro-ecological analyses of the topology of these networks have revealed common properties across them, such as truncated power-law degree distributions [@jordano1987] or nestedness [@bascompte2003]. In addition, this type of large scale analyses can help understanding landscape level processes that cannot be explored at the community level, such as ecological patterns across geographic regions [@olesen2002; @traveset2016] or environmental gradients [@ramos2010; @rech2016; @saunders2023]. Although macro-ecological approaches make significant contributions to knowledge, they tend to be rare, and are strongly influenced by the spatio-temporal availability and nature of the data [@burkle2011; @trojelsgaard2013]. For instance, plant-pollinator studies tend to differ in sampling effort and methodology which can impact the structure of the resulting plant-pollinator networks [@gibson2011; @jordano2016]. Further, despite little sampling effort could capture most of the relevant functional species [@hegland2010], most plant-pollinator networks have unobserved interactions because undersampling [@olesen2011; @chacoff2012]. Thus, the lack of strong spatio-temporal coverage, along with the current intrinsic limitations of sampling plant-pollinator networks, highlight the need to keep gathering and integrating informative species interaction data to properly unravel the different ecological processes that shape plant-pollinator interactions at large scales.

@@

Bees are responsible for the majority of the interactions at the metaweb level, but their relative relevance changed across habitats and bioclimatic regions. For instance, plant-pollinator communities in the Mediterranean were dominated by bees, while communities in Alpine or Boreal regions were fly species rich or fly-dominated. These patterns are consistent with our current understanding of bee diversity, which peaks in dry or temperate areas [@orr2021; @leclercq2023]; and with the fact that colder environments (i.e., altitude and latitude wise) harbor a larger fraction of fly pollinators compared to other taxa [@elberling1999; @lefebvre2018]. In addition, beetles were only common in the Mediterranean region. Although the networks from the database are visitation networks and do not capture pollinator efficiency [@ballantyne2015], the high proportion of beetles as floral visitors in the Mediterranean provides further support for their potential role as pollinators in this region [@herrera2019; @leon2022]. The number of butterfly species and interactions were relatively low compared to the other taxa. While Europe contains fewer butterfly species than other regions of the world [@ollerton2017], their relevance as pollinators is likely underestimated within this database. This is because a large fraction of studies (~40%) did not sample butterflies, and conventional sampling methods for monitoring other insect pollinators (e.g., bees or flies) may be inadequate for sampling plant-butterfly interactions [@isaac2011]. On the contrary, honey bees were present in `r round(apis_percent_networks, 1)`% of networks and conducted on average a third of the total interactions per network. Indeed, this proportion is higher than the one found when considering only natural communities at a global scale [~13%\; @hung2018], which highlights their key role as potential pollinators of many flowering plants. However, the rise of beekeeping in Europe has led to high honey bee densities, which can negatively impact wild pollinators and the plants they visit [@steffan2000; @magrach2017; @herrera2020], emphasising the need to evaluate these impacts and implement actions if necessary.
